Kalman tracking and estimation of continuous gravitational waves with a pulsar timing array

### 2023-08-30 docs/manuscript/paper.tex
PP plot at canonical low h

 If we consider the median values of the one-dimensional posteriors we can see that for some parameters the median is almost identically equal to the injected value (e.g. $\Omega)

 The clear exception is $\iota$ which exhibits a notable bias, and lies outside the with the 90\% credible interval, with the posterior median shifted from the injected value by $\sim 0.3$ radians.

 For a single realisation of the noise it is not clear if this discrepancy is a systematic effect, i.e. a bias in the parameter estimate, or else just a random variation specific to this particular noise realisation. We explore this further in Section \ref{sec:multiple_noise} and Section \ref{sec:bias} and show that this is a bias that results from dropping the pulsar terms, similar to the bias reported in \cite{Zhupulsarterms}.

  $, \Phi_0, \delta$) whereas for others the median has a slight discrepancy from the injected value (e.g. $\psi, \alpha$), whilst still being contained within the credible interval.

  This potential bias will also be investigated alongside the bias in $\iota$ in Section \ref{sec:multiple_noise} and Section \ref{sec:bias}.

 Similar results are derived for the the 3$N$ parameters of $\boldsymbol{\theta}_{\rm psr, reduced}$ and the parameters are generally recovered unambiguously, but we do not display the resulting posterior distributions here since the inference of $\boldsymbol{\theta}_{\rm gw}$ is the main focus of this work. We are able to simultaneously estimate both $\boldsymbol{\theta}_{\rm gw}$ and $\boldsymbol{\theta}_{\rm psr, reduced}$.

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{images/likelihood_surface}
	\caption{Likelihood $\log \mathcal{L}$ (Equation \ref{eq:likelihood}) surface across the $\iota - h_0$ parameter space for a single noise realisation. The red point labels the location of the likelihood maxima. The likelihood surface has been normalised with respect to the absolute value of the maxima. A strong likelihood ridge is evident in the parameter space where the Kalman filter produces similar likelihood values, making it difficult for sampling algorithms to converge to a consistent value.}
	\label{fig:likelihood_surface}
\end{figure}
%\begin{figure*}
%	\centering
%	\includegraphics[width=\textwidth]{images/compare_iotas_lines2}
%	\caption{One-dimensional marginalized posteriors for $\iota$ over 9 realisations of the noise where $h$ is (a) an unknown free parameter (left panel) and (b) fixed at its true value $= 10^{-12}$ (right panel). The thick vertical orange line labels the true injected value. The dashed vertical lines denote the median of the corresponding coloured posterior. Fixing $h$ reduces the variance between different noise realisations but results in a clear bias between the true and inferred values of $\iota$. The left hand panel is equivalent to the marginalized posteriors for $\iota$ shown in Figure \ref{fig:corner_plot_2} (4th row, 4th column), but with the $(0.16,0.84)$ quantiles replaced with the $0.5$ quantile.}
%	\label{fig:iota}
%\end{figure*}
The above discussion was for a single realisation of the noise processes. It is important to confirm that our method is robust against different realisations of the noise and that the specific realisation of the noisy data used in Sections \ref{sec:parameter_estim} and \ref{sec:detection} is not particularly advantageous or convenient for our method. In this Section we confirm that we are able to infer comparable results for different noise realisations. To this end we take our representative example from Table \ref{tab:parameters_and_priors} and generate 1000 realisations of the process noise and measurement noise. For each realisation of the data we can then independently estimate the static parameters, $\boldsymbol{\theta}$. In Figure \ref{fig:corner_plot_2} we show the results for the  estimated values of $\boldsymbol{\theta}_{\rm gw}$ for nine realisations of the noise. This figure is analogous to the single noise realisation of Figure \ref{fig:corner_plot_1}. We plot only 9 realisations of the noise rather than the full set of 1000 realisations so as to not overcrowd the figure. As in Section \ref{sec:parameter_estim} the estimates of $\boldsymbol{\theta}_{\rm psr}$ are recovered unambiguously across the 1000 noise realisations and for brevity we do not show the results here. \newline

@@

To this end we use the the Wasserstein distance \citep[WD,][]{Wasserstein,Villani2009} from optimal transport theory which defines an intuitive notion of similarity between probability distributions. The WD is the cost of an optimal strategy for moving probability mass between two distributions from position $x$ to position $y$, with respect to some cost function $c(x,y)$. Throughout this work we use the first moment of the Wasserstein distance which for one-dimensional distributions $\mu$ and  $\nu$ is defined
\begin{eqnarray}
	W_1(\mu,\nu)=  \inf_{\lambda \in \Gamma(\mu, \nu)} \int |x-y| d \lambda (x,y)
\end{eqnarray}
where $\lambda(x,y)$ is the transport plan, $\Gamma(\mu, \nu)$ is the set of all couplings of $\mu$ and $\nu$ and $ |x-y|$ defines the cost function. The WD has an additional useful physical interpretation given by the Monge-Kantorovich duality:
\begin{eqnarray}
	| \boldsymbol{E}(X_{\mu} )-\boldsymbol{E}(Y_{\nu} ) | \leq W_1(\mu, \nu) \label{eq:WDdefn}
\end{eqnarray}
where $X_{\mu}$ and $Y_{\nu}$ are random variables of the distributions $\mu$ and $\nu$ respectively, and $\boldsymbol{E}$ denotes the expected value. We can see that the WD bounds the difference in the expected values. For example, if we compare two posteriors for $\iota$ calculated for two different noise realisations, a WD = 1.0 rad tells us that the difference in the expected value of $\iota$ from the two distributions is less than 1 rad. Beyond optimal transport theory, the WD is a popular metric in machine learning \citep{2017arXiv170107875A}, climate modelling \citep{2022JCli...35.1215P,2023QJRMS.149..843K}, computational biology \citep{GONZALEZDELGADO2023168053} and geophysics \citep{2023GeoRL..5003880M}. A full review of the Wasserstein distance, which may be unfamiliar to some readers, is presented in Appendix \ref{sec:wasserstein} \textcolor{red}{TK: is this too much info on Wasserstein? Maybe just say "its a metric between probability distributions" and move the rest to appendix so as not to interrupt main astrophysics arguments.}

>>>>

 For some of the parameters the estimated posterior is approximately symmetric about the true injected value (e.g. $\Omega$). In this case the posterior median and the injected value are approximately coincident. For other parameters (e.g. $\iota$) the distribution is not symmetric about the injected value and the posterior median and the injected value are not coincident. The median value of the posterior for $\iota$ is shifted by $\sim 0.35$ rad relative to the injected value, although the injected value does still remain inside the 90\% credible interval. Similar effects are seen, albeit with a smaller shift, in other parameters such as $\delta$ and $\alpha$. For a single realisation of the noise it is not clear if this discrepancy is a systematic effect, i.e. a bias in the parameter estimate, or else just a random variation specific to this particular noise realisation. We explore this further in Section \ref{sec:multiple_noise} and Section \ref{sec:bias} and show that this is a bias that results from dropping the pulsar terms, similar to the bias reported in \cite{Zhupulsarterms}. Similar results are derived for the the 3$N$ parameters of $\boldsymbol{\theta}_{\rm psr, reduced}$ and the parameters are generally recovered unambiguously, but we do not display the resulting posterior distributions here since the inference of $\boldsymbol{\theta}_{\rm gw}$ is the main focus of this work. We are able to simultaneously estimate both $\boldsymbol{\theta}_{\rm gw}$ and $\boldsymbol{\theta}_{\rm psr, reduced}$.

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{images/likelihood_surface}
	\caption{Likelihood $\log \mathcal{L}$ (Equation \ref{eq:likelihood}) surface across the $\iota - h_0$ parameter space for a single noise realisation. The red point labels the location of the likelihood maxima. The likelihood surface has been normalised with respect to the absolute value of the maxima. A strong likelihood ridge is evident in the parameter space where the Kalman filter produces similar likelihood values, making it difficult for sampling algorithms to converge to a consistent value.}
	\label{fig:likelihood_surface}
\end{figure}
%\begin{figure*}
%	\centering
%	\includegraphics[width=\textwidth]{images/compare_iotas_lines2}
%	\caption{One-dimensional marginalized posteriors for $\iota$ over 9 realisations of the noise where $h$ is (a) an unknown free parameter (left panel) and (b) fixed at its true value $= 10^{-12}$ (right panel). The thick vertical orange line labels the true injected value. The dashed vertical lines denote the median of the corresponding coloured posterior. Fixing $h$ reduces the variance between different noise realisations but results in a clear bias between the true and inferred values of $\iota$. The left hand panel is equivalent to the marginalized posteriors for $\iota$ shown in Figure \ref{fig:corner_plot_2} (4th row, 4th column), but with the $(0.16,0.84)$ quantiles replaced with the $0.5$ quantile.}
%	\label{fig:iota}
%\end{figure*}
The above discussion was for a single realisation of the noise processes. It is important to confirm that our method is robust against different realisations of the noise and that the specific realisation of the noisy data used in Sections \ref{sec:parameter_estim} is not particularly advantageous or convenient for our method. In this Section we confirm that we are able to infer comparable results for different noise realisations. To this end we take our representative example from Table \ref{tab:parameters_and_priors} and generate 1000 realisations of the process noise and measurement noise. For each realisation of the data we can then independently estimate the static parameters, $\boldsymbol{\theta}$. In Figure \ref{fig:corner_plot_2} we show the results for the  estimated values of $\boldsymbol{\theta}_{\rm gw}$ for nine realisations of the noise. This figure is analogous to the single noise realisation of Figure \ref{fig:corner_plot_1}. We plot only 9 realisations of the noise rather than the full set of 1000 realisations so as to not overcrowd the figure. As in Section \ref{sec:parameter_estim} the estimates of $\boldsymbol{\theta}_{\rm psr}$ are recovered unambiguously across the 1000 noise realisations and for brevity we do not show the results here. \newline

@@

To this end we use the the Wasserstein distance \citep[WD;][]{Wasserstein,Villani2009} from optimal transport theory which defines an intuitive notion of similarity between probability distributions. The WD is the cost of an optimal strategy for moving probability mass between two distributions from position $x$ to position $y$, with respect to some cost function $c(x,y)$. Throughout this work we use the first moment of the Wasserstein distance which for one-dimensional distributions $\mu$ and  $\nu$ is defined
\begin{eqnarray}
	W_1(\mu,\nu)=  \inf_{\lambda \in \Gamma(\mu, \nu)} \int |x-y| d \lambda (x,y)
\end{eqnarray}
where $\lambda(x,y)$ is the transport plan, $\Gamma(\mu, \nu)$ is the set of all couplings of $\mu$ and $\nu$ and $ |x-y|$ defines the cost function. The WD has an additional useful physical interpretation given by the Monge-Kantorovich duality:
\begin{eqnarray}
	| \boldsymbol{E}(X_{\mu} )-\boldsymbol{E}(Y_{\nu} ) | \leq W_1(\mu, \nu) \label{eq:WDdefn}
\end{eqnarray}
where $X_{\mu}$ and $Y_{\nu}$ are random variables of the distributions $\mu$ and $\nu$ respectively, and $\boldsymbol{E}$ denotes the expected value. We can see that the WD bounds the difference in the expected values. For example, if we compare two posteriors for $\iota$ calculated for two different noise realisations, a WD = 1.0 rad tells us that the difference in the expected value of $\iota$ from the two distributions is less than 1 rad. Beyond optimal transport theory, the WD is a popular metric in machine learning \citep{2017arXiv170107875A}, climate modelling \citep{2022JCli...35.1215P,2023QJRMS.149..843K}, computational biology \citep{GONZALEZDELGADO2023168053} and geophysics \citep{2023GeoRL..5003880M}. A full review of the Wasserstein distance, which may be unfamiliar to some readers, is presented in Appendix \ref{sec:wasserstein} \textcolor{red}{TK: is this too much info on Wasserstein? Maybe just say "its a metric between probability distributions" and move the rest to appendix so as not to interrupt main astrophysics arguments.}
